
First and foremost, I thank my God for all the blessings in my life.

Then, I would like to thank my wonderful loving husband Ali who has always been for me in my rough and good times. Ali became my family and best friend in my life away from home. He believed in me and did all he could to let me achieve my goals. I thank him for all support, understanding and encouragements. He is the one secret behind every success I make or challenge I overcome. 

I thank my two adorable sons Hassan and Rida for being in my life - making it joyful, exciting and a bit more busy! 

I thank my loving family: Mom, Dad, Sisters and Brothers. I appreciate that they are always standing up for me during all of my milestones, pushing me hard, trusting me, and cheering for every success I make. My special thanks to my great super Mom. She is the one person who believes in me the most. I thank her for your prayers, love and constant communication regardless of the time difference and long distance. I do what I do to make her proud. 

I thank my great advisor Professor Eric Walkingshaw. I thank his high confidence in me, his inspiring words throughout my journey and his powerful encouragement which kept me going. I honor his time, lessons, feedback and enthusiasm. I appreciate his understanding and accommodations for my special needs and restricted schedule. I appreciate his respect to my values. I give him all the credit for me continuing my studies and accomplishing this work.

I thank Professor Martin Erwig who has been a main source of knowledge in the PL group. I appreciate everything I learned from him through classes or the reading group.

My very special thanks to my mentor Jed Irvine for the great sessions in improving my technical writing. He has effectively improved my skills, reviewed my paper, and gave me the time and effort I needed with great generosity.

I thank my PL research mates for the reading group meetings where we exchanged many great ideas. I specially thank my group mates Jeff, Parisa, Fariba and Qiaoran. I appreciate all of their feedback during the writing workshops. 

I thank the great faculty members of OSU who I learned from or worked with. I thank Professor Prasad Tadepalli who I enjoyed working with on the GTA assignment project. I thank Professor Jennifer Parham-Mocello who has always shown love and support. I thank the adviser Calvin Hughes for his unlimited advising and support.

I thank Professor Glencora Borradaile for her support and empathy. I will absolutely miss my many stops by her office. She has always helped me through my difficult times in personal and graduate life.

I thank my very special friend Mahsa for her exceptional support, especially after moving out from Corvallis and starting to commute. I thank her for always checking on me, hosting me in your apartment during lunch time and cooking for me. In my hardest days during pregnancy and commuting, Mahsa has been there for me.

I thank my friends in Corvallis and Portland. I appreciate all of the great times we spend together. I thank my very special extended family who always provide me with love and support.

I thank OSU for the GTA position which provided me with teaching skills and secured me financially. I also thank SACM for the scholarship which provided financial stability plus child care support.

I thank my committee members---Eric Walkingshaw, Martin Erwig, Arash Termehchy, and Vergara Hector for their time, willingness to evaluate my work, and feedback.